Response time of a plasmonic distributed feedback laser in large-signal modulation regime

### 2019-05-07 DFB_Pulse_LaTeX/DFB_pulse_inProcess.tex
Second revision, final version

For a quantitative description of the temporal dynamics of a plasmonic DFB laser under pulse pumping, we consider a laser based on an Au film perforated by a periodic array of nanosized holes and coated with the layer of the active medium, see Fig.~\ref{fig1}. We consider that the thickness of the Au film is $100$ $\text{nm}$, the diameter of the holes is $160$ $\text{nm}$, and the distance between holes is $470$ $\text{nm}$. The surface of the entire periodic plasmonic structure is $190 \times 190$ $\mu \text{m}^2$. We consider that the transition frequency of the active medium, $\omega_{\sigma}$, is close to the frequency of the second Bloch condition in the plasmonic structure, i.e. $k_\text{pl}(\omega_{\sigma}) \approx G$, where $k_\text{pl}$ is a wavevector of the mode in the plasmonic structure and $G$ is a reciprocal lattice vector \cite{TennerJOpt,TennerACSPhot,nefedkin2018acsphot}. The linewidth of the active medium is about $100 \text{nm}$. Similar plasmonic DFB lasers have been implemented in a number of experiments \cite{BeijnumPRL,TennerJOpt,TennerACSPhot} and their parameters have been well studied \cite{BeijnumPRL,TennerJOpt,TennerACSPhot,nefedkin2018acsphot}.

@@

The existence of a critical size of the pump beam can be explained with the example of a single-mode model. To justify the single-mode approximation, we calculate the distribution of the electric field intensity over the area of the plasmonic DFB laser at different moments of time, see Fig.~\ref{fig3}.
In our simulations we take into account several hundred of the Bloch modes of the plasmonic structure, which together form the EM field distribution in the active medium. One can see that this distribution does not change in the first and second stages of the system response. In other words, there is a dynamical balance between relaxation processes in the plasmonic structure, the propagation of the EM field outside the pumped area and the stimulated amplification of the EM field inside the pumped area. When the diameter of the pump beam is larger than the critical value, $d_{\text{beam}} > d_{\text{beam}}^*$, the electric field is located inside the pumped region of the active medium, see Figs.~\ref{fig3}(d)--(f).
In the opposite case, when $d_{\text{beam}} < d_{\text{beam}}^*$, the main part of the electric field is outside the pumped area, see Figs.~\ref{fig3}(a)--(c).
However, in both cases, the spatial distribution of the electric field intensity remains almost unchanged during its temporal evolution.

We use the fact of the constancy of the EM field distribution to construct a single-mode model. To do this, we assume that there is a single collective mode, the EM field distribution of which coincides with the EM field distribution found in the numerical simulations of the multimode problem. Furthermore, we assume that the pumped active medium interacts only with such a single collective mode. Within these approximations, the system dynamics can be described by the following system of equations \cite{SiegmanLasers}:
\begin{gather} 
\dot n =  - 2{\gamma _a}n + 2{N_{\text{at}}}{\Omega _R}\varphi \label{ZE1}
\\
\dot \varphi = - \left( {{\gamma _a} + {\gamma _\sigma }} \right)\varphi + {\Omega _R}\left( {2nD + D + 1} \right)/2 \label{ZE2}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 4{\Omega _R}\varphi, \label{ZE3}
\end{gather}
where $n$ is the number of photons in the collective mode, $D$ is the mean population inversion of pumped atoms of the active medium, $\varphi$ describes the mean energy flow from the active medium to the collective mode, and $N_{\text{at}}$ is the number of atoms of the active medium lying in the collective mode.
We suppose that the relaxation rate of the EM field in the collective mode is equal to the characteristic relaxation rate of the modes of the plasmonic structure, i.e., $\gamma_a \simeq 10^{12} \text{s}^{-1}$.
Note that the finite size of the pump beam can lead to the appearance of edge effects and influence on the mode quality factor, see \cite{hakala2017lasing}.
In the single-mode model, we do not take into account these effects.
The typical relaxation rate of the population inversion of the active medium is $\gamma_D \simeq 10^9 \text{s}^{-1}$ while the phase relaxation rate is $\gamma_{\sigma}\simeq 10^{13} \text{s}^{-1}$.
Thus, the phase relaxation rate is much larger than other relaxation rates, $\gamma_\sigma \gg \gamma_a, \gamma_D$.
For this reason, one can adiabatically eliminate $\varphi$ from Eqs.~(\ref{ZE1})--(\ref{ZE3}) and obtain the conventional rate equations
\begin{gather} 
\dot n =  - 2 {\gamma _a}n + {N_{\text{at}}}\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE1}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 2\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE2}
\end{gather}
As initial condition we take $D(0) = 1$ and $n(0) = 0$, i.e., all atoms are inverted and there are no photons in the mode.
The temporal dynamics of such a system is well-known \cite{SiegmanLasers}; it is shown in Fig.~\ref{fig4}.

@@

The second stage corresponds to the time when the exponential term in Eq.~(\ref{inv0}) is dominant, and the dynamics is strongly nonlinear.
This nonlinear term describes the stimulated energy flow from the atoms to the collective mode.
In this case, from Eq.~(\ref{inv0}) we obtain the equation
\begin{equation} 
\dot D =  - {\gamma _D}\frac{{2\xi \eta D}}{{\left( {\xi  - 1} \right)}} \exp \left[\left( {\xi  - 1} \right) 2 {\gamma _a}t \right] \label{invnl1}
\end{equation}
which has the solution
\begin{equation} 
D(t) \simeq \exp \left(  - \frac{{\xi \eta {\gamma _D}}}{{{\left( {\xi  - 1} \right)}^2}{\gamma _a}}  \exp \left[\left( {\xi  - 1} \right) 2 {\gamma _a}t \right] \right) = \exp \left(  - \frac{{{\xi ^2}}}{{{N_{\text{at}}}{{\left( {\xi  - 1} \right)}^2}}}\exp \left[\left( {\xi  - 1} \right) 2 {\gamma _a}t \right] \right).
\end{equation}
The characteristic time for $D$ to change is 
\begin{equation} 
{t_{delay}} = (2\gamma _a)^{ - 1}{\left( {\xi  - 1} \right)^{ - 1}}\ln \left( {{N_{\text{at}}}{{\left( {\xi - 1} \right)}^2}/{\xi ^2}} \right). \label{tdele}
\end{equation}
At this time, a sharp decrease of $D$ appears, see Fig.~\ref{fig4}b.
Note that the assumption that the evolution of $D$ is much slower than the evolution of $n$ is valid when $t_{delay} \gg \gamma_a^{-1} \xi^{-1}$, i.e., $\ln \left( {{N_{\text{at}}}{{\left( {\xi  - 1} \right)}^2}/{\xi ^2}} \right) \gg 1$.
When $\xi \gg 1$, the assumption is correct if $N_{\text{at}} \gg 1$, which is true for the considered system.

@@

In consequense, there is the diameter of the pump beam, $d_{\text{beam}}^*$, necessary to achieve the short response time. This diameter is about the decay length of the EM field in the plasmonic structure and for the considered plasmonic laser is $\sim 15 \mu$m. The existence of the critical pump beam size gives the opportunity to reduce the energy consumption because further increase of the pump beam size does not shorten the response time.
In turn, if the area of the pump beam is small, there is the opportunity to reduce the size of the device.
Thus, the obtained results pave the way to creating ultrafast optoelectronic devices based on plasmonic lasers with low power consumption which have the potential of on-chip integration.

>>>>

For a quantitative description of the temporal dynamics of a plasmonic DFB laser under pulse pumping, we consider a laser based on an Au film perforated by a periodic array of nanosized holes and coated with the layer of the active medium, see Fig.~\ref{fig1}. We consider that the thickness of the Au film is $100$ $\text{nm}$, the diameter of the holes is $160$ $\text{nm}$, and the distance between holes is $470$ $\text{nm}$. The surface of the entire periodic plasmonic structure is $190 \times 190$ $\mu \text{m}^2$. We consider that the transition frequency of the active medium, $\omega_{\sigma}$, is close to the frequency of the second Bloch condition in the plasmonic structure, i.e. $k_\text{pl}(\omega_{\sigma}) \approx G$, where $k_\text{pl}$ is a wavevector of the mode in the plasmonic structure and $G$ is a reciprocal lattice vector \cite{TennerJOpt,TennerACSPhot,nefedkin2018acsphot}. The linewidth of the active medium is about $100$ $\text{nm}$. Similar plasmonic DFB lasers have been implemented in a number of experiments \cite{BeijnumPRL,TennerJOpt,TennerACSPhot} and their parameters have been well studied \cite{BeijnumPRL,TennerJOpt,TennerACSPhot,nefedkin2018acsphot}.

@@

The existence of a critical size of the pump beam can be explained with the example of a single-mode model. To justify the single-mode approximation, we calculate the distribution of the electric field intensity over the area of the plasmonic DFB laser at different moments of time, see Fig.~\ref{fig3}.
In Eqs. (\ref{eq1})-(\ref{eq3}), we take into account several hundred of the Bloch modes of the plasmonic structure, which together form the EM field distribution in the active medium. Numerical simulations show that this distribution does not change in the first and second stages of the system response. In other words, there is a dynamical balance between relaxation processes in the plasmonic structure, the propagation of the EM field outside the pumped area and the stimulated amplification of the EM field inside the pumped area. When the diameter of the pump beam is larger than the critical value, $d_{\text{beam}} > d_{\text{beam}}^*$, the electric field is located inside the pumped region of the active medium, see Figs.~\ref{fig3}(d)--(f).
In the opposite case, when $d_{\text{beam}} < d_{\text{beam}}^*$, the main part of the electric field is outside the pumped area, see Figs.~\ref{fig3}(a)--(c).
However, in both cases, the spatial distribution of the electric field intensity remains almost unchanged during its temporal evolution.

We use the fact of the constancy of the EM field distribution to construct a single-mode model. To do this, we assume that there is a single collective mode, the EM field distribution of which coincides with the EM field distribution found in the numerical simulations of the multimode problem. Furthermore, we assume that the pumped active medium interacts only with such a single collective mode. Within these approximations, the system dynamics can be described by the following system of equations \cite{SiegmanLasers}:
\begin{gather} 
\dot n =  - 2{\gamma _a}n + 2{N_{\text{at}}}{\Omega _R}\varphi \label{ZE1}
\\
\dot \varphi = - \left( {{\gamma _a} + {\gamma _\sigma }} \right)\varphi + {\Omega _R}\left( {2nD + D + 1} \right)/2 \label{ZE2}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 4{\Omega _R}\varphi, \label{ZE3}
\end{gather}
where $n$ is the number of photons in the collective mode, $D$ is the mean population inversion of pumped atoms of the active medium, $\varphi$ describes the mean energy flow from the active medium to the collective mode, and $N_{\text{at}}$ is the number of atoms of the active medium lying in the collective mode.
We suppose that the relaxation rate of the EM field in the collective mode is equal to the characteristic relaxation rate of the modes of the plasmonic structure, i.e., $\gamma_a \simeq 10^{12} \text{s}^{-1}$.
Note that the finite size of the pump beam can lead to the appearance of edge effects and influence on the mode quality factor, see \cite{hakala2017lasing}.
In the single-mode model, we do not take into account these effects for simplicity.
We will further see that this simple model is in good agreement with the numerical simulations of Eqs. (\ref{eq1})-(\ref{eq3}).
The typical relaxation rate of the population inversion of the active medium is $\gamma_D \simeq 10^9 \text{s}^{-1}$ while the phase relaxation rate is $\gamma_{\sigma}\simeq 10^{13} \text{s}^{-1}$.
Thus, the phase relaxation rate is much larger than other relaxation rates, $\gamma_\sigma \gg \gamma_a, \gamma_D$.
For this reason, one can adiabatically eliminate $\varphi$ from Eqs.~(\ref{ZE1})--(\ref{ZE3}) and obtain the conventional rate equations
\begin{gather} 
\dot n =  - 2 {\gamma _a}n + {N_{\text{at}}}\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE1}
\\
\dot D =  - {\gamma _D}\left( {D + 1} \right) - 2\Omega _R^2\left( {2nD + D + 1} \right)/{\gamma _\sigma } \label{RE2}
\end{gather}
As initial condition we take $D(0) = 1$ and $n(0) = 0$, i.e., all atoms are inverted and there are no photons in the mode.
The temporal dynamics of such a system is well-known \cite{SiegmanLasers}; it is shown in Fig.~\ref{fig4}.

@@

The second stage corresponds to the time when the exponential term in Eq.~(\ref{inv0}) is dominant, and the dynamics is strongly nonlinear.
This nonlinear term describes the stimulated energy flow from the atoms to the collective mode.
In this case, from Eq.~(\ref{inv0}) we obtain the equation
\begin{equation} 
\dot D =  - {\gamma _D}\frac{{2\xi \eta D}}{{\left( {\xi  - 1} \right)}} \exp \left[\left( {\xi  - 1} \right) 2 {\gamma _a}t \right] \label{invnl1}
\end{equation}
which has the solution
\begin{equation} 
D(t) \simeq \exp \left(  - \frac{{\xi \eta {\gamma _D}}}{{{\left( {\xi  - 1} \right)}^2}{\gamma _a}}  \exp \left[\left( {\xi  - 1} \right) 2 {\gamma _a}t \right] \right) = \exp \left(  - \frac{{{\xi ^2}}}{{{N_{\text{at}}}{{\left( {\xi  - 1} \right)}^2}}}\exp \left[\left( {\xi  - 1} \right) 2 {\gamma _a}t \right] \right).
\end{equation}
The characteristic time for $D$ to change is 
\begin{equation} 
{t_\text{delay}} = (2\gamma _a)^{ - 1}{\left( {\xi  - 1} \right)^{ - 1}}\ln \left( {{N_{\text{at}}}{{\left( {\xi - 1} \right)}^2}/{\xi ^2}} \right). \label{tdele}
\end{equation}
At this time, a sharp decrease of $D$ appears, see Fig.~\ref{fig4}b.
Note that the assumption that the evolution of $D$ is much slower than the evolution of $n$ is valid when $t_\text{delay} \gg \gamma_a^{-1} \xi^{-1}$, i.e., $\ln \left( {{N_{\text{at}}}{{\left( {\xi  - 1} \right)}^2}/{\xi ^2}} \right) \gg 1$.
When $\xi \gg 1$, the assumption is correct if $N_{\text{at}} \gg 1$, which is true for the considered system.

@@

To conclude we have shown that there is the diameter of the pump beam, $d_{\text{beam}}^*$, necessary to achieve the short response time. This diameter is about the decay length of the EM field in the plasmonic structure and for the considered plasmonic laser is $\sim 15 \mu$m. The existence of the critical pump beam size gives the opportunity to reduce the energy consumption because further increase of the pump beam size does not shorten the response time.
In turn, if the area of the pump beam is small, there is the opportunity to reduce the size of the device.
Thus, the obtained results pave the way to creating ultrafast optoelectronic devices based on plasmonic lasers with low power consumption which have the potential of on-chip integration.
